---
title: "FlaringABM input validation"
author: "Nick Willems"
---

```{r}
library(ggplot2)
library(data.table)
```

# Lease expenses
## Capital costs  
Gas leases should have only one well
```{r}
print(leases_emp[OIL_GAS_CODE=="G", summary(N)])
```

Expect a capital expenditures to be about [$7/BOE](zotero://select/items/0_BSNIWDAF)
```{r}
leases_emp[, cat("All leases:", sum(capEx) / sum(EUR), "\t Assigned leases:", .SD[!is.na(firmID), sum(capEx) / sum(EUR)])]
```

## Operating costs per MCF of gas at oil and gas leases  
```{r fig.cap="Gas lease operating expenditures are skewed left (cheaper) while oil leases are skewed right"}
ggplot(leases[gas_MCF+csgd_MCF>0]) +
    geom_density(aes(x=opEx_pMCF + ifelse(csgd_MCF==0, 0, capEx_csgd / csgd_MCF / lifetime), color=OIL_GAS_CODE)) +
    labs(color="Oil or Gas lease?", x="Average Cost per MCF") +
    theme(legend.position = c(0.85, 0.9)) +
    xlim(0, 5)
```

## Operating costs per BBL of oil at oil leases  
```{r}
ggplot(leases[oil_BBL>0]) +
    geom_density(data= function(x) {x[,"area":= ifelse(area=="Spraberry", "Midland Basin", area)]},
                aes(x=cost_oil/oil_BBL, color=area)) +
    scale_x_log10() +
    labs(x="$ / barrel", main="Operating expenses per barrel of oil")
```

| Region            | Year  | Mean  | Min   | Max   |
| ---               | ---   | ---   | ---   | ---   |
| Eagle Ford        | 2017  | 29    | 10    | 50    |
| Eagle Ford        | 2021  | 16.58 | 7     | 35    |
| Permian - Delaware| 2017  | 33    | 15    | 55    |
| Permian - Delaware| 2021  | 26.19 | 7     | 50    |
| Permian - Midland | 2017  | 24    | 10    | 55    |
| Permian - Midland | 2021  | 27.05 | 6     | 60    |
Table: Break-even operating expenses ($/barrel) reported by Dallas Fed
([2017](https://www.dallasfed.org/research/surveys/des/2017/1701.aspx#tab-questions),
[2021](https://www.dallasfed.org/research/surveys/des/2021/2101.aspx#tab-questions)):  

# Lease assignment summary  
### Number of assigned leases   
```{r}
print(leases[is.na(firmID), .N, by=OIL_GAS_CODE])
```

### Number of leases left to be discovered  
```{r}
print(leases[!is.na(firmID), .N, by=OIL_GAS_CODE])
```

### Lifetime estimation
Goodness of fit among historical leases
```{r}
leases_emp[(expiration<201912) & (start>199301),
            .("R2"= summary(lm(log(m_rem)~log(q_i/q_curve)))$r.squared), by=area]
```

```{r}
ggplot(leases_emp, aes(x=q_i/EUR, y=lifetime)) +
    geom_point(aes(color=ifelse((expiration<201912) & (start>199301),"Observed","Predicted")), alpha=0.3) +
    scale_x_log10() + scale_y_log10() +
    theme_bw() + theme(plot.title=element_text(hjust=0.5)) +
    facet_grid(.~area) +
    labs(title="Lease lifetimes based on production profile", color="")
```

### Expiring & new production
```{r}
ggplot(leases[!is.na(firmID), sum(oil_BBL + cond_BBL), by=.("ex"=lifetime + t_found - Params$t0)]) +
    geom_col(aes(x=ex, y=V1)) +
    geom_hline(aes(yintercept=median(V1, na.rm=TRUE)), lty=2) +
    geom_text(data=function(x) {x[,.(median(V1, na.rm=TRUE))]},
         aes(y=V1, label=sprintf("Median: %.02f", V1)), x=-Inf, hjust=-0.01, vjust=-1) +
    labs(title="Quantity of expiring oil production", y="Oil + Condensate (BBL)", x="Time (Months)") +
    coord_cartesian(xlim=c(0, Params$tf - Params$t0))
```

```{r}
cat("Potential new production\n")
print(leases[is.na(firmID), summary(oil_BBL + cond_BBL)])
```

### Percent of operators with no gas leases  
```{r}
# Empirical
print(market_shares[, sum(scaled_gas_MCF==0)/.N])
# Generated
print(leases[, sum(gas_MCF), by=firmID][, sum(V1==0)/.N])
```

# Distributions of operator production volumes
```{r}
par(mfrow=c(1,2))
hist(log(market_shares$scaled_oil_BBL), breaks="scott", freq=FALSE,
    main="Empirical", xlab="log(Oil (BBL))", xlim=c(0,20), ylim=c(0,0.2))
market_shares[, {curve(dnorm(x, mean(log(scaled_oil_BBL)), sd(log(scaled_oil_BBL))), col="blue", add=TRUE); NULL}]
hist(log(leases[!is.na(firmID), sum(oil_BBL), by=firmID]$V1), breaks="scott", freq=FALSE,
    main="Generated", xlab="log(Oil (BBL))", xlim=c(0,20), ylim=c(0,0.2))
market_shares[, {curve(dnorm(x, mean(log(scaled_oil_BBL)), sd(log(scaled_oil_BBL))), col="blue", add=TRUE); NULL}]
```

```{r}
par(mfrow=c(1,2))
hist(log(market_shares$scaled_gas_MCF), breaks="scott", freq=FALSE,
    main="Empirical", xlab="log(Gas (MCF))", xlim=c(0,29), ylim=c(0,0.2))
market_shares[scaled_gas_MCF>0, .(log(scaled_gas_MCF))][, {curve(dnorm(x, mean(V1), sd(V1)), col="blue", add=TRUE); NULL}]
hist(log(leases[!is.na(firmID), sum(gas_MCF), by=firmID]$V1), breaks="scott", freq=FALSE,
    main="Generated", xlab="log(Gas (MCF))", xlim=c(0,20), ylim=c(0,0.2))
market_shares[scaled_gas_MCF>0, .(log(scaled_gas_MCF))][, {curve(dnorm(x, mean(V1), sd(V1)), col="blue", add=TRUE); NULL}]
```

Joint density of oil and gas outputs
```{r}
ggplot(market_shares, aes(x=scaled_oil_BBL, y=scaled_gas_MCF)) +
    stat_density_2d(geom="raster", aes(fill=after_stat(density)), contour=FALSE) +
    scale_fill_viridis_c() +
    geom_point(data=leases[!is.na(firmID), .(sum(oil_BBL), sum(gas_MCF)), by=firmID][V1>0 & V2>0],
                aes(x=V1, y=V2, color="red"), alpha=0.8, shape=1, size=2) +
    scale_color_manual(labels="Generated", values="red") +
    scale_x_log10() +
    scale_y_log10() +
    labs(title="Joint distribution of firm oil and gas production", x="Oil (BBL)", y="Gas (MCF)") +
    labs(color="", fill="Emprical")
```

Joint density of oil and casinghead gas outputs
```{r}
ggplot(market_shares, aes(x=scaled_oil_BBL, y=scaled_csgd_MCF)) +
    stat_density_2d(geom="raster", aes(fill=after_stat(density)), contour=FALSE) +
    scale_fill_viridis_c() +
    geom_point(data=leases[!is.na(firmID), .(sum(oil_BBL), sum(csgd_MCF)), by=firmID][V1>0 & V2>0],
                aes(x=V1, y=V2, color="red"), alpha=0.8, shape=1, size=2) +
    scale_color_manual(labels="Generated", values="red") +
    scale_x_log10() +
    scale_y_log10() +
    labs(title="Joint distribution of firm oil and casinghead gas production", x="Oil (BBL)", y="Gas (MCF)") +
    labs(color="", fill="Emprical")
```


# Market conditions
## Oil
### Supply curve
```{r}
ggplot(leases[!is.na(firmID) & oil_BBL>0][order(opEx_pBBL)]) +
    geom_line(aes(x=cumsum(oil_BBL), y=opEx_pBBL)) +
    scale_y_log10() +
    labs(x="Q", y="P", title="Oil supply curve")
```

## Natural gas supply and demand
### Supply curve
```{r}
ggplot(leases[class=="developed"][order(opEx_pMCF)]) +
    geom_line(aes(x=cumsum(gas_MCF), y=opEx_pMCF, color="Only un-associated gas")) +
    geom_line(aes(x=cumsum(csgd_MCF + gas_MCF), y=opEx_pMCF, color="Including casinghead gas")) +
    theme(legend.position=c(0.2, 0.8), legend.background=element_blank()) +
    labs(x="Q (MCF / month)", y="Marginal Cost ($)", title="Natural gas supply curve", color="")
```

### Demand curve
```{r fig.cap="Sample demand curves using single months. Model demand curves use samples of one or more months"}
ggplot(demand$historical_market, aes(color=factor(month, levels=month.abb))) +
    geom_point(aes(x=q, y=p), shape=4) +
    geom_line(data= function(x) {x[sample(.N, 5), {"shift"= q - (q_TX*frac*0.95);
                                                    cbind(shift, demand$new_schedule(0, sample_set=.BY)(x$q - shift))},
                                    by=.(year, month)][p_grey>0]},
            aes(x=q+shift, y=p_grey, group=interaction(year, month)), show.legend=FALSE) +
    geom_text(aes(q, p, label=substr(year,3,4)), size=3, hjust=0) +
    labs(title="Sampling of representative demand curves", x="Quantity (MCF)", y="Price ($/MCF)", color="month")
```

### Emergent market
```{r}
ggplot(leases[!is.na(firmID)][order(opEx_pMCF)]) +
    geom_line(aes(x=cumsum(gas_MCF), y=opEx_pMCF, color="Only un-associated gas")) +
    geom_line(aes(x=cumsum(csgd_MCF + gas_MCF), y=opEx_pMCF, color="Including casinghead gas"),
                data=leases[class=="developed"][order(opEx_pMCF)]) +
    geom_line(aes(x=cumsum(csgd_MCF + gas_MCF), y=opEx_pMCF, color="Assuming 100%\ncasinghead gas capture")) +
    geom_line(data= function(x) {rbindlist(lapply(1:5, function(z) demand$new_schedule(0)(x[, cumsum(gas_MCF+csgd_MCF)])), idcol=TRUE)},
                aes(x=q, y=p_grey, group=.id, color="Sample demand")) +
    theme(legend.position=c(0.85, 0.15), legend.background=element_blank()) +
    ylim(0,7) +
    labs(x="Q (MCF / month)", y="Price ($)", title="Natural gas market", color="")
```

```{r}
ggplot(leases[firms[behavior!="flaring"], on="firmID"][order(opEx_pMCF)]) +
    geom_line(aes(x=cumsum(gas_MCF), y=opEx_pMCF, color="Only un-associated gas")) +
    geom_line(aes(x=cumsum(csgd_MCF + gas_MCF), y=opEx_pMCF, color="Assuming 100%\ncasinghead gas capture")) +
    geom_line(data= function(x) {rbindlist(lapply(1:5, function(z) demand$new_schedule(0.1)(x[, cumsum(gas_MCF+csgd_MCF)])), idcol=TRUE)},
                aes(x=q, y=p_green, group=.id, color="Sample green demand")) +
    theme(legend.position=c(0.85, 0.15), legend.background=element_blank()) +
    ylim(0,7) +
    labs(x="Q (MCF / month)", y="Price ($)", title="Green natural gas market", color="")
```
