---
title: "FlaringABM firms characteristics"
author: "Nick Willems"
output:
  html_document:
    code_folding: hide
  pdf_document:
    keep_tex: true
knit: (function(input, ...) { rmarkdown::render(input, output_format='all', ...) })
---

```{css, echo=FALSE, eval=knitr::is_html_output()}
/* avoid displaying title and author in html output */
.title{
    display: none;
}
.author{
    display: none;
}
```

```{r, echo=FALSE, eval=knitr::is_latex_output()}
library(formatR)
knitr::opts_chunk$set(tidy=TRUE,
                    tidy.opts=list(width.cutoff=40),
                    fig.path=sub("/", "_tex/", knitr::opts_chunk$get('fig.path')))
```

```{r, echo=FALSE, include=FALSE}
library(knitr)
library(printr)
# define a method for objects of the class data.table
registerS3method("knit_print", "data.table", function(x, ...) {
    if (!shouldPrint(x)) {
        return(invisible())
    } else {
        NextMethod("knit_print") # print as data.frame
    }
})
```

CHARACTERIZE UPSTREAM OIL AND GAS FIRMS WHO PRODUCE GAS ALONGSIDE OIL USING TEXAS RAILROAD COMMISSION DATA.  


```{r}
library(data.table)
library(ggplot2)
library(readxl)
```

# Monthly production from new wells
EIA Drilling productivity report 
[Source](https://www.eia.gov/petroleum/drilling/archive/2021/10/)  
```{r}
EIA_dpr <- rbindlist(lapply(list("Eagle Ford"="Eagle Ford Region", "Permian"="Permian Region"),
                            function(x) read_xlsx("./data/dpr-data.xlsx", sheet=x, skip=1)), idcol="region")
setnames(EIA_dpr, "Month", "date")

EIA_dpr[, "new_BBL":= `Rig count` * `Production per rig...3`]
EIA_dpr[, "retired_BBL":= -`Legacy production change...4`]
EIA_dpr[, "total_BBL":= cumsum(new_BBL - retired_BBL) + first(`Total production...5`), by=region]

ggplot(EIA_dpr, aes(x=date)) +
    geom_line(aes(y=total_BBL, color="Calculated")) +
    geom_line(aes(y=`Total production...5`, color="Provided")) +
    facet_grid(region~., scale="free_y") +
    labs(title="Calculated and provided total oil production", y="Production (BBL)")
```

```{r}
ggplot(EIA_dpr, aes(x=date)) +
    geom_line(aes(y=new_BBL / retired_BBL, color=region)) +
    geom_hline(aes(yintercept=median(new_BBL / retired_BBL, na.rm=TRUE)), lty=2) +
    geom_text(data=function(x) {x[,.(median(new_BBL / retired_BBL, na.rm=TRUE))]},
        aes(y=V1, label=sprintf("Median: %.02f",V1)), x=-Inf, hjust=-1, vjust=-1) +
    labs(title="Oil production gained and lost from new and retiring wells", y="Production gained / Production lost")
```

```{r}
ggplot(EIA_dpr, aes(x=date)) +
    geom_line(aes(y=new_BBL / retired_BBL, color=region)) +
    geom_hline(aes(yintercept=median(new_BBL / retired_BBL, na.rm=TRUE)), lty=2) +
    geom_text(data=function(x) {x[,.(median(new_BBL / retired_BBL, na.rm=TRUE))]},
        aes(y=V1, label=sprintf("Median: %.02f",V1)), x=-Inf, hjust=-1, vjust=-1) +
    labs(title="Oil production gained and lost from new and retiring wells", y="Production gained / Production lost") +
    ylim(0,3)
```

Output the compiled drilling productivity data  
```{r}
EIA_dpr[, "new_MCF":= `Rig count` * `Production per rig...6`]
EIA_dpr[, "retired_MCF":= -`Legacy production change...7`]

fwrite(EIA_dpr[!is.na(new_BBL), .(date, region, new_BBL, retired_BBL, new_MCF, retired_MCF)], "./processed/EIA_dpr.csv")
```

# Monthly county production 
From TX RRC Production Data Query Dump 
 [Source](https://mft.rrc.texas.gov/link/fe3298a2-8788-4234-b2a0-90ee41558d75),
 [Manual](https://portalvhdskzlfb8q9lqr9.blob.core.windows.net/media/47731/pdq-dump-user-manual.pdf)
```{r}
OG_COUNTY_CYCLE <- fread("unzip -p ./data/PDQ_DSV.zip OG_COUNTY_CYCLE_DATA_TABLE.dsv", sep="}", quote="",
                        select=list(character=c("DISTRICT_NAME"), numeric=c("COUNTY_NO","CYCLE_YEAR","CYCLE_MONTH"),
                                    numeric=c("CNTY_OIL_PROD_VOL","CNTY_GAS_PROD_VOL","CNTY_CSGD_PROD_VOL")))

# which counties are represented in the subset of leases
OG_COUNTY_CYCLE[, "subset":= (COUNTY_NO %in% unique(fread("./processed/wells.csv")$API_COUNTY_CODE))]

# based on TXRRC descriptions of the Permian and Eagle Ford
OG_COUNTY_CYCLE[subset==TRUE &
                # [Permian](https://www.rrc.texas.gov/oil-and-gas/major-oil-and-gas-formations/permian-basin/)
                !((DISTRICT_NAME %in% c("7C","08","8A")) |
                # [Eagle Ford](https://www.rrc.texas.gov/oil-and-gas/major-oil-and-gas-formations/eagle-ford-shale/)
                (COUNTY_NO %in% c(13, 21, 25, 41, 51, 123, 127, 149, 163, 177, 185, 255, 283, 285,
                                287, 289, 297, 311, 313, 323, 331, 395, 471, 477, 479, 493, 507))),
                "subset":= FALSE]

# What proportion of gas comes from areas where we have lease data
print(OG_COUNTY_CYCLE[CYCLE_YEAR>=2010 & CYCLE_YEAR<2021,
                lapply(.SD, function(x) sum(x[subset==TRUE]) / sum(x)),
                by=.(CYCLE_MONTH, CYCLE_YEAR), .SDcols=patterns("PROD_VOL$")])
```

Summary of production in Texas and the subset of areas with lease data
```{r}
lease_prod <- OG_COUNTY_CYCLE[CYCLE_YEAR>=2010 & CYCLE_YEAR<2021,
                            .("oil_total"= sum(CNTY_OIL_PROD_VOL), "oil_subset"= sum(CNTY_OIL_PROD_VOL[subset]),
                            "gas_total"= sum(CNTY_GAS_PROD_VOL), "gas_subset"= sum(CNTY_GAS_PROD_VOL[subset]),
                            "csgd_total"= sum(CNTY_CSGD_PROD_VOL), "csgd_subset"= sum(CNTY_CSGD_PROD_VOL[subset])),
                    by=.(CYCLE_MONTH, CYCLE_YEAR)]

lease_prod[, "Date":= as.Date(paste("01", CYCLE_MONTH, CYCLE_YEAR), format='%d %m %Y')]

ggplot(lease_prod, aes(x=Date)) +
    geom_point(aes(y=gas_subset/gas_total, color="gas")) +
    geom_point(aes(y=csgd_subset/csgd_total, color="csgd")) +
    geom_point(aes(y=(gas_subset+csgd_subset)/(gas_total+csgd_total), color="gas+csgd")) +
    labs(title="Production from counties with lease data", y="% of Texas", color="")

summary(lease_prod[, (gas_subset+csgd_subset)/(gas_total+csgd_total)])
```

# Monthly operator production 
From TX RRC Production Data Query Dump
 [Source](https://mft.rrc.texas.gov/link/fe3298a2-8788-4234-b2a0-90ee41558d75),
 [Manual](https://portalvhdskzlfb8q9lqr9.blob.core.windows.net/media/47731/pdq-dump-user-manual.pdf)
```{r}
OG_OPERATOR_CYCLE <- fread("unzip -p ./data/PDQ_DSV.zip OG_OPERATOR_CYCLE_DATA_TABLE.dsv", sep="}", quote="",
                            colClasses=list(numeric=c("OPER_OIL_PROD_VOL","OPER_GAS_PROD_VOL","OPER_CSGD_PROD_VOL")))
```

## Types of operators based on mean yearly production
```{r}
# production from 2010 through 2020
operator_types <- OG_OPERATOR_CYCLE[CYCLE_YEAR>=2010 & CYCLE_YEAR<2021,
                    lapply(.SD, mean), .SDcols=c("OPER_OIL_PROD_VOL","OPER_GAS_PROD_VOL","OPER_CSGD_PROD_VOL"),
                    by=.(OPERATOR_NO, CYCLE_YEAR)]

# active operators
operator_types <- operator_types[!(OPER_OIL_PROD_VOL==0 & OPER_GAS_PROD_VOL==0 & OPER_CSGD_PROD_VOL==0)]

# determine which operators are producing gas alongside oil
operator_types[(OPER_OIL_PROD_VOL>0) & (OPER_GAS_PROD_VOL==0), "firm_type":= "oil_only"]
operator_types[(OPER_OIL_PROD_VOL==0) & (OPER_GAS_PROD_VOL>0), "firm_type":= "gas_only"]
operator_types[(OPER_OIL_PROD_VOL>0) & (OPER_CSGD_PROD_VOL>0), "firm_type":= "associated"]

OG_OPERATOR_CYCLE[operator_types, on=c("OPERATOR_NO","CYCLE_YEAR"), "firm_type":= firm_type]
```

```{r}
# number of operators
OG_OPERATOR_CYCLE[firm_type=="oil_only", .("N"=uniqueN(OPERATOR_NO)), keyby=CYCLE_YEAR]
OG_OPERATOR_CYCLE[firm_type=="gas_only", .("N"=uniqueN(OPERATOR_NO)), keyby=CYCLE_YEAR]
OG_OPERATOR_CYCLE[firm_type=="associated", .("N"=uniqueN(OPERATOR_NO)), keyby=CYCLE_YEAR]

# percent of total gas production coming from 'associated' operators
OG_OPERATOR_CYCLE[CYCLE_YEAR>=2010 & CYCLE_YEAR<2021,
                lapply(.(rowSums(.SD)), function(x) sum(x[firm_type=="associated"], na.rm=TRUE) / sum(x)),
                by=CYCLE_YEAR, .SDcols=c("OPER_GAS_PROD_VOL","OPER_CSGD_PROD_VOL")]
```

## Characterize active, 'associated' operators
```{r}
operators <- copy(OG_OPERATOR_CYCLE[firm_type=="associated"])

operators[lease_prod, on=c("CYCLE_MONTH", "CYCLE_YEAR"),
    c("frac_oil", "frac_gas", "frac_csgd"):= .(oil_subset/oil_total, gas_subset/gas_total, csgd_subset/csgd_total)]

# average production scaled according to how much production comes from areas corresponding to lease data
operators[, "scaled_oil_BBL":= median(OPER_OIL_PROD_VOL * frac_oil), by=OPERATOR_NO]
operators[, "scaled_gas_MCF":= median(OPER_GAS_PROD_VOL * frac_gas), by=OPERATOR_NO]
operators[, "scaled_csgd_MCF":= median(OPER_CSGD_PROD_VOL * frac_csgd), by=OPERATOR_NO]

# subset operators who produce above a threshold
operators <- operators[scaled_oil_BBL > 0]
operators <- operators[(scaled_gas_MCF + scaled_csgd_MCF) > 0]
```

```{r, fig.cap="Oil production is nearly log-normally distributed", results='hide'}
operators[, .(log(scaled_oil_BBL))][,
            {"h_data"= hist(V1, breaks="scott", plot=FALSE);
            plot(exp(h_data$mids), h_data$density, log='x', type='h', lwd=3, lend=2,
            main="Distribution of firm scaled oil production", xlab="Oil (BBL)", ylab="", ylim=c(0,0.25));
            curve(dnorm(log(x), mean(V1), sd(V1)), lwd=2, col='blue', add=TRUE); NULL}]
```

```{r}
ggplot(operators[OPER_OIL_PROD_VOL>0], aes(log(OPER_OIL_PROD_VOL))) +
    geom_histogram(aes(y=after_stat(density)), bins=nclass.scott(log(operators[OPER_OIL_PROD_VOL>0]$OPER_OIL_PROD_VOL))) +
    stat_function(fun=dnorm, args=operators[, .(log(scaled_oil_BBL))][, .("mean"=mean(V1), "sd"=sd(V1))]) +
    facet_wrap(.~CYCLE_YEAR)
```

```{r}
operators[, sum(scaled_oil_BBL), by=.(CYCLE_YEAR, CYCLE_MONTH)][, as.list(summary(V1)), keyby=CYCLE_YEAR]
```

```{r, fig.cap="Gas production is nearly log-normally distributed", results='hide'}
operators[scaled_gas_MCF>0, .(log(scaled_gas_MCF))][,
            {"h_data"= hist(V1, breaks="scott", plot=FALSE);
            plot(exp(h_data$mids), h_data$density, log='x', type='h', lwd=3, lend=2,
            main="Distribution of firm scaled gas production", xlab="Gas (MCF)", ylab="", ylim=c(0,0.25));
            curve(dnorm(log(x), mean(V1), sd(V1)), lwd=2, col='blue', add=TRUE); NULL}]
```

```{r}
operators[, sum(scaled_gas_MCF), by=.(CYCLE_YEAR, CYCLE_MONTH)][, as.list(summary(V1)), keyby=CYCLE_YEAR]
```

```{r, fig.cap="Casinghead gas production is not log-normally distributed", results='hide'}
operators[scaled_csgd_MCF>0, .(log(scaled_csgd_MCF))][,
            {"h_data"= hist(V1, breaks="scott", plot=FALSE);
            plot(exp(h_data$mids), h_data$density, log='x', type='h', lwd=3, lend=2,
            main="Distribution of firm scaled casinghead gas production", xlab="Gas (MCF)", ylab="", ylim=c(0,0.25));
            curve(dnorm(log(x), mean(V1), sd(V1)), lwd=2, col='blue', add=TRUE); NULL}]
```

Output operators
```{r}
fwrite(operators, "./processed/firm_market_shares.csv")
```

# Financial data  
```{r}
# CRSP/Compustat Merged (Annual)
CCM <- fread("./data/CRSPCompustat Merged - Fundamentals Annual.csv.gz", key=c("fyear","LPERMCO","conm"))

# SIC major group 1311: oil and gas extraction- Crude Petroleum and Natural Gas
# NAIC group 2111 Oil and Gas Extraction
CCM_og <- CCM[(sic==1311 | naics==2111) & !is.na(sale) & sale>=0]

# Number of firms
ggplot(CCM_og[, .("Number of firms"=uniqueN(LPERMCO)), by=fyear]) +
    geom_col(aes(x=fyear, y=`Number of firms`))

CCM_og[fyear>2010, .("Number of firms"=uniqueN(LPERMCO)), by=fyear]
```

### Calculate characteristics using firm data
Described in [Leary & Roberts 2014](zotero://select/items/0_2I87I24H)
```{r}
CCM_og[, "market_share":= sale / sum(sale), by=fyear] # compare to market share
# market value
CCM_og[, "market_value":= mkvalt] # compare to market value
# capital investment
CCM_og[, "CI":= c(NA, capx[-1] / ppent[-.N]), by=LPERMCO] # compare to cash
CCM_og[CI==Inf, "CI":= NA]
```

```{r}
ggplot(CCM_og[fyear>=2010]) +
    geom_density(aes(x=CI)) +
    scale_x_log10() +
    facet_wrap(.~fyear)
```

Output financial data
```{r}
fwrite(CCM_og[fyear>=2010], "./processed/firm_finances.csv")
```
