---
title: "FlaringABM market history"
author: "Nick Willems"
---

CHARACTERIZE WELL COST INFLATION.  


```{r}
library(data.table)
library(readxl)
library(ggplot2)
```

# Upstream cost index  
[MIT: The future of geothermal energy](zotero://select/items/0_FHEHLW9V) (pg 6-13 & 6-34) 
presents an oil & gas well cost index that only includes onshore wells. 
[EIA](https://www.eia.gov/dnav/ng/NG_ENR_WELLCOST_S1_A.htm) presents a cost index with more historical data, 
but it is not clear whether offshore wells are also included. 
```{r}
MIT_index <- fread("./data/MIT_composite_index.csv")
MIT_index[, "source":= "MIT"]

EIA_index <- read_xls("./data/NG_ENR_WELLCOST_S1_A.xls", sheet="Data 1", skip=2)
setDT(EIA_index)
EIA_index[, "source":= "EIA"]
EIA_index[, "year":= as.numeric(substr(Date, 1, 4))]

EIA_index[, "nom_index_1977":= 100 * .SD / .SD[year==1977][[1]],
    .SDcols="U.S. Nominal Cost per Crude Oil, Natural Gas, and Dry Well Drilled (Thousand Dollars per Well)"]

well_cost_index <- rbind(EIA_index[!between(year, 1976, 2004), .(year, nom_index_1977, source)],
                        MIT_index[year>=1976])[order(year)]

well_cost_index <- rbind(EIA_index[, .(year, nom_index_1977, source)], MIT_index)[order(year)]

well_cost_index[, "composite":= (source=="MIT" & year>=1976) | (source=="EIA" & !between(year, 1976, 2004))]
```

```{r echo = T, results = 'hide', fig.cap="Very strong agreement between EIA & MIT cost indices prior to 1992"}
color_scheme <- c("Composite"="black", "EIA"="red", "MIT"="blue")
plot(nom_index_1977 ~ year, well_cost_index[composite==TRUE], type="b", col=color_scheme[1], cex=1.3,
    main="Nominal cost index\n(1977=100)", xlab="Year", ylab="Index (1977=100)")
lapply(unique(well_cost_index$source), function(x)
    points(nom_index_1977 ~ year, well_cost_index[source==x], type="b", col=color_scheme[x], pch="+"))
grid()
legend("topleft", names(color_scheme), col=color_scheme, pch=c(1,3,3), title="Index Source")
```

```{r}
well_cost_index[EIA_index, on="year",
    "scaler":=  `U.S. Real Cost per Crude Oil, Natural Gas, and Dry Well Drilled (Thousand Dollars per Well)` /
                `U.S. Nominal Cost per Crude Oil, Natural Gas, and Dry Well Drilled (Thousand Dollars per Well)`]
well_cost_index[, "real_index_1977":= nom_index_1977 * (scaler / scaler[year==1977]), by=source]

well_cost_index[, "real_index_2005":= real_index_1977 / real_index_1977[year==2005]]
```

[IEA](https://www.iea.org/data-and-statistics/charts/iea-us-shale-upstream-cost-index-2005-2020)
have a third upstream cost index for US shale.  
[US EIA](zotero://select/items/0_EJYISQT4) & [IEA](zotero://select/items/0_RAVQRI9Y)
draw similar conclusions about historical cost data (e.g., 2015 costs were ~30% lower than 2012 values).
```{r}
IEA_index <- fread(text=   "year, real_index_2005
                            2005, 100
                            2006, 118.7
                            2007, 128.5
                            2008, 141.2
                            2009, 108
                            2010, 125.1
                            2011, 141.6
                            2012, 133.7
                            2013, 130.4
                            2014, 130.1
                            2015, 97.5
                            2016, 83.4
                            2017, 90.5
                            2018, 101.2
                            2019, 99.7")

IEA_index[, "real_index_2005":= real_index_2005/100]
IEA_index[, "source":= "IEA"]
```

```{r}
well_cost_index <- rbind(well_cost_index, IEA_index, fill=TRUE)[order(year)]
well_cost_index[, "composite":= (source=="MIT" & year>=1976) |
                                (source=="EIA" & year<1976) |
                                (source=="IEA")]
```

```{r echo = T, results = 'hide', fig.cap="Close agreement between EIA & IEA cost indices for 2005, 2006"}
color_scheme <- c(color_scheme, c("IEA"="orange"))
plot(real_index_2005 ~ year, well_cost_index[composite==TRUE], type="b", col=color_scheme[1], cex=1.3, ylim=c(0,2.3),
    main="Real cost index", xlab="Year", ylab="Index (2005=1)")
lapply(unique(well_cost_index$source), function(x)
    points(real_index_2005 ~ year, well_cost_index[source==x], type="b", col=color_scheme[x], pch="+"))
grid()
legend("topleft", names(color_scheme), col=color_scheme, pch=c(1,3,3,3), title="Index Source")
```

Output the composite 2014 index  
```{r}
well_cost_index[, "real_index_2014":= real_index_2005 / real_index_2005[year==2014]]

fwrite(well_cost_index[composite==TRUE, .(year, source, real_index_2014)], "./processed/well_cost_index.csv")
```

